Spatiotemporal Changes in Netrin/Dscam1 Signaling Dictate Axonal Projection Direction in Drosophila Small Ventral Lateral Clock Neurons

Version 1

Abstract

Abstract

Axon projection is a spatial and temporal-specific process in which the growth cone receives environmental signals guiding axons to their final destination. However, the mechanisms underlying changes in axonal projection direction without well-defined landmarks remain elusive. Here, we present evidence showcasing the dynamic nature of axonal projections in Drosophila’s small ventral lateral clock neurons (s-LNvs). Our findings reveal that these axons undergo an initial vertical projection in the early larval stage, followed by a subsequent transition to a horizontal projection in the early-to-mid third instar larvae. The vertical projection of s-LNv axons correlates with mushroom body calyx expansion, while the s-LNv-expressed Down syndrome cell adhesion molecule (Dscam1) interacts with Netrins to regulate the horizontal projection. During a specific temporal window, locally newborn dorsal clock neurons (DNs) secrete Netrins, facilitating the transition of axonal projection direction in s-LNvs. Our study establishes a compelling in vivo model to probe the mechanisms of axonal projection direction switching in the absence of clear landmarks. These findings underscore the significance of dynamic local microenvironments in the synergetic regulation of axonal projection direction transitions.

Introduction

During nervous system development, neurons extend axons to reach their targets in order to build functional neural circuits (Cang & Feldheim, 2013; Kaas, 1997). The growth cone, which is specialized at the tip of the extending axon, is critical for receiving multiple guidance signals from the external environment to guide axon projection (Lowery & Vactor, 2009; Mortimer et al., 2008; Vitriol & Zheng, 2012). Guidance cues regulate cytoskeletal dynamics through growth cone-specific receptors, steering axons via attractive or repulsive signals (Araújo & Tear, 2003; Koch et al., 2012; Kolodkin & Tessier-Lavigne, 2011; Esther T Stoeckli, 2018; Zang et al., 2021). Over the last 30 years, several guidance cues have been identified and divided into four classical families: Semaphorins, Ephrins, Netrins, and Slits (Derijck et al., 2010; Dickson, 2002; Goodhill, 2016; X.-T. Li et al., 2014; Orioli & Klein, 1997). The emergence of other kinds of guidance cues, including morphogens, growth factors, and cell adhesion molecules, increases the complexity of axon pathfinding (Chan & Odde, 2008; Chédotal et al., 2012; Kahn & Baas, 2016; Vitriol & Zheng, 2012).

When going through a long-distance pathfinding process in the complex and highly dynamic in vivo environment, axons frequently undergo multiple changes in their projection directions (Squarzoni et al., 2015). Crucially, intermediate targets play a pivotal role by providing vital guiding information that enables axons to transition their projection directions and embark upon the subsequent stages of their journey, ultimately leading them to their final destination (de Ramon Francàs et al., 2017; Garel & Rubenstein, 2004; Squarzoni et al., 2015). Certain intermediate targets can respond to the growth cone’s signaling to maintain a dynamic balance between attractive and repulsive forces(Bron et al., 2007; Martins et al., 2022). One of the well-studied intermediate target projection types is the midline cross model (Derijck et al., 2010; Yang et al., 2009). As a clear landmark, axons need to project to the midline first and then leave it immediately (Gorla & Bashaw, 2020; Neuhaus-Follini & Bashaw, 2015). The midline floor plate cells secreted long-range attractive signals to attract commissural axons towards them (Kennedy et al., 1994; Serafini et al., 1996). Once there, these cells emit signals of rejection to prevent excessive closeness of axons, thus aiding the axon in departing the floor plate to reach its intended destination (Kidd et al., 1998; Long et al., 2004; E T Stoeckli et al., 1997). Comparatively, the typical axonal projection process within the in vivo milieu undergoes multiple transitions in projection direction without a distinct landmark, such as the midline (Garel & López-Bendito, 2014; Gezelius & López-Bendito, 2017). However, how the axons change their projection direction without well-defined landmarks is still unclear.

Drosophila small ventral lateral clock neurons (s-LNvs) exhibit the typical tangential projection pattern in the central brain (Agrawal & Hardin, 2016; Gummadova et al., 2009; Hardin, 2017; Helfrich-Förster, 1997). The s-LNv axons originate from the ventrolateral soma and project to the dorsolateral area of the brain, and subsequently undergo a axon projection direction shift, projecting horizontally toward the midline with a short dorsal extension (Helfrich-Förster et al., 2007). Several studies have reported the impact of various factors, including the Slit-Robo signal (Oliva et al., 2016), Lar (leukocyte-antigen-related) (Agrawal & Hardin, 2016), dfmr1 (Okray et al., 2015), dTip60 (Pirooznia et al., 2012), and Dscam1 long 3’ UTR (Zhang et al., 2019), on the axon outgrowth of s-LNvs in the adult flies’ dorsal projection. However, the mechanism underlying the projection direction switch of s-LNv axons remains largely unknown. Here, we show that s-LNvs and newborn dorsal clock neurons (DNs) generate spatiotemporal-specific guidance cues to precisely regulate the transition of projection direction in s-LNv axons. This regulation uncovers unexpected interdependencies between axons and local microenvironment dynamics during the navigation process, enabling accurate control over axon projection.

Results

s-LNvs axons change their projection direction in the early-to-mid third instar larvae

The stereotypical trajectory of the s-LNvs axon projection can be succinctly characterized as an initial vertical extension originating from the ventrolateral brain, followed by a directional pivot at the dorsolateral protocerebrum, ultimately leading to a horizontal projection towards the midline. Immunoreactivity for pigment-dispersing factor (PDF) (Cyran et al., 2005), representing the pattern of LNv neurons, is initially detected in the brains of 1st-instar larvae 4-5 hours after larval hatching (ALH) (Helfrich-Förster, 1997). To elucidate the intricate axon pathfinding process during development, we employed a dual-copy Pdf-GAL4 to drive the expression of double-copy UAS-mCD8::GFP, visualizing the s-LNvs at the larval stage (Figure 1A). At 8 hours ALH, the GFP signal strongly co-localized with the PDF signal within the axons. This remarkable co-localization continued throughout the larval stage, particularly concentrated at the axon’s terminal end (Figure 1A).

s-LNvs axon projection dynamics during the larval stage.

(A) Images showing the growth process of s-LNvs during larval development. Larval brains were stained with anti-PDF (white) and HRP (magenta) antibodies. Different time points indicate the hours after larval hatching (ALH).

(B) Left panel: schematic diagram illustrating the method used to measure the degree of vertical projection. One hemisphere of the larvae brain is depicted. larval neuropil (gray), s-LNvs (green), optic lobe (blue). Right panel: graphs showing the average vertical projection length at different developmental stages, presented as mean ± SD.

(C) Left panel: schematic diagram illustrating the method used to measure the degree of horizontal projection. Right panel: graphs showing the average horizontal projection (A.U.: arbitrary unit) at different developmental stages, presented as mean ± SD. The red line segment indicates the stage at which the axonal projection undergoes a directional transition.

(D) Schematic representation of s-LNvs vertical to horizontal projection directional shift during 72-96 hours ALH. One hemisphere of the larvae brain (gray), larval neuropil (white), s-LNvs (green).

For (B), (C), 8 h (n = 7), 24 h (n = 10), 48 h (n = 14), 72 h (n = 16), 96 h (n = 13), 120 h (n = 16).

We then quantified the process of s-LNvs axonal projection. The vertical projection distance of the s-LNvs axon was ascertained by measuring the vertical span between the lowest point of the optical lobe and the highest point of the axon’s projection terminal. Statistical analyses revealed a steadfast linear increase in the vertical projection distance of the s-LNvs axon as larval development progressed (Figure 1A-B). To determine the horizontal projection distance of the s-LNvs axon, we applied the methodology employed by Olive et al. for measuring adult flies (Oliva et al., 2016). A tangent line is drawn precisely at the pivotal juncture where the s-LNvs axon undergoes a shift in its projection direction. We define the Arbitrary Unit (A.U.) as the ratio between the horizontal distances from this point to both the end of the axon projection and the midline (Figure 1A and 1C). Remarkably, at the time points of 8, 24, 48, and 72 hours ALH, the A.U. value for s-LNvs axon projection in the horizontal direction remained relatively stable at approximately 0.13. However, at 96 and 120 hours ALH, a significant and striking increase in the A.U. value ensued, eventually reaching approximately 0.25 (Figure 1C). These findings suggest that the precise transition of s-LNv axon projection from vertical towards the horizontal direction occurs in the early-to-mid third instar larvae, specifically during 72 to 96 hours ALH (Figure 1D).

Vertical projection of s-LNvs axons correlate with mushroom body calyx expansion

The axons of s-LNv project dorsally, coming in close proximity to the dendritic tree of the mushroom body (MB), specifically the calyx (Figures 2A and 2C). This spatial arrangement facilitates potential interactions between the s-LNvs axon and the MB. MB predominantly comprises 2500 intrinsic neurons, known as Kenyon Cells (KCs) (Puñal et al., 2021). The development of the MB is a dynamic process characterized by three primary types of KCs: α/β, α’/β’, and γ, each following distinct temporal schedules of birth. The γ cells emerge from the embryonic stage until early 3rd-instar larval phase, while the α’/β’ cells are born during the latter half of larval life, and the α/β cells appear in the post-larval stage (Lee et al., 1999; Puñal et al., 2021). We noticed that s-LNv axons exhibit only vertical growth before early 3rd-instar larval stage, which coincides with the emergence of γ KCs. Consequently, we have contemplated the potential correlation between the vertical projection of s-LNv axons and the growth of the mushroom body.

Development of vertical projection process alongside Mushroom Body growth.

(A) Top: Spatial position relationship between calyx and s-LNvs labeled by OK107-GAL4, which labels MB γ and α’β’ neurons, at different developmental time points. Bottom: Spatial position relationship between calyx and s-LNvs labeled by Tab2-201Y-GAL4, which labels MB γ neurons, Larvae brains were stained with anti-PDF (green) and GFP (magenta) antibodies. Different time indicated hours ALH. The white dotted line indicates the calyx region.

(B) Graphs showing the mushroom body calyx’s area and vertical projection length averaged over different development stage are presented as mean ± SD. OK107: 24 h (n = 10), 48 h (n = 8), 72 h (n = 12), 96 h (n = 12), 120 h (n = 9). Pearson’s r = 0.9987, P < 0.0001. Tab2-201Y: 24 h (n = 6), 48 h (n = 12), 72 h (n = 13), 96 h (n = 13), 120 h (n = 11). Pearson’s r = 0.9963, P = 0.0003.

(C) Schematic representation of synergetic development of s-LNvs vertical projection and mushroom body calyx. Mushroom body (purple), s-LNvs (Green).

(D) Images of mushroom body ablation in the developing larval stages. Larvae brains were stained with anti-PDF (white) and HRP (magenta) antibodies. Different time indicated hours ALH.

(E) Quantification of vertical projection length in Control (Tab2-201Y >GFP) and Ablation (Tab2-201Y >GFP,rpr,hid) flies. Data are presented as mean ± SD. Control: 24 h (n = 6), 48 h (n = 14), 72 h (n = 13), Ablation: 24 h (n = 6), 48 h (n = 12), 72 h (n = 12). Two-tailed Student’s t tests were used. ns, p > 0.05; ****p < 0.0001.

Next, we closely monitored the spatial relationship between the calyx and the axon projections of the s-LNvs throughout different larval developmental stages (Figure 2A). We used OK107-GAL4 and Tab2-201Y-GAL4 to drive the expression of mCD8::GFP in all KC subtypes or specifically in γ KCs (Pauls et al., 2010), respectively, to visualize MB KCs. As development progressed, the calyx area defined by both GAL4 drivers exhibited a consistent expansion. Statistical analyses further confirmed that the calyx area increased proportionally with the advancement of developmental time, demonstrating a linear relationship (Figure 2B-C). Additionally, Pearson correlation analysis unveiled a robust positive correlation between the calyx area and the vertical projection distance of s-LNvs axon (r = 0.9987, P < 0.0001 for OK107-GAL4; r = 0.9963, P = 0.0003 for Tab2-201Y-GAL4) (Figure 2B-C). These compelling observations suggest that the MB calyx, and potentially specifically the calyx of γ KCs, exert significant influence on the vertical projection of the s-LNvs axon.

To validate our idea, we employed a straightforward approach to evaluate any changes in s-LNvs axon projection after ablating γ KCs using the apoptosis-promoting factors RPRC and HID driven by the Tab2-201Y GAL4. Initially, we confirmed the effectiveness of our approach. In the control group, the GFP signal in γ KCs remained unchanged, while the ablation group showed a significant reduction or complete absence of the signal (Figure 2-figure supplement 1). Subsequently, we examined the axon projection of s-LNvs under these conditions. In the non-ablated group, the vertical distance increased linearly during development, following the expected pattern (Figure 2D-E). However, in the ablation group, although the initial vertical projection of s-LNvs appeared normal, there was a noticeable decrease in vertical projection distance at 48 hours ALH. This vertical development trend ceased thereafter, with the vertical projection distance measuring only 20 to 30 µm at 72 hours ALH (Figure 2D-E). Importantly, in later larval stages, we observed that s-LNvs in certain larvae became undetectable (data not shown), suggesting the potential occurrence of apoptosis in these neurons following MB ablation. These results underscore a significant association between MB development, specifically the γ KCs, and the vertical projection dynamics of s-LNv axons.

s-LNv-expressed Dscam1 mediate s-LNv axons horizontal projection

To decipher the signaling pathways governing s-LNvs axon horizontal projection, we performed a screening with a total of 285 targets by using the RNA-interfering (RNAi) approach (Supplementary File 3). To optimize efficiency, we chose to use Clk856-GAL4, which exhibits high expression in s-LNv during early larval stages, despite also being expressed in subsets of other clock neurons (Gummadova et al., 2009). Significantly, the knockdown of Down syndrome cell adhesion molecule (Dscam1) exhibited a marked reduction in the horizontal projection distance of s-LNv axons (Figure 3-figure supplement 1A-B). To validate the role of s-LNv-expressed Dscam1 in governing axon horizontal projection, we employed the Pdf-GAL4, which specifically targeted s-LNvs during the larval stage. At 96 hours ALH, vertical projection showed no significant changes in Pdf-GAL4;UAS-Dscam1-RNAi flies, but a noticeable deficit was observed in the horizontal projection of the s-LNvs axon (Figure 3A-C). These observations strongly indicate that Dscam1 specifically regulates the horizontal projection of s-LNvs axon.

Critical role of Dscam1 in horizontal projection.

(A) Images of s-LNvs in Pdf >GFP and Pdf >GFP, Dscam1RNAi fly. White line segment represents the horizontal projection distance of s-LNvs. Larvae brains were stained with anti-PDF (white) and HRP (magenta) antibodies at 96 hours ALH. Pdf >GFP: (n = 13), Pdf >GFP, Dscam1RNAi: (n = 15).

(B) Quantification of vertical projection length in Pdf >GFP and Pdf >GFP, Dscam1RNAi flies.

Data are presented as mean ± SD. Two-tailed Student’s t tests, ns, p > 0.05.

(C) Quantification of horizontal A.U. in Pdf >GFP and Pdf >GFP, Dscam1RNAi flies. Data are presented as mean ± SD. Two-tailed Student’s t tests, ****p < 0.0001.

(D) Images of Dscam1 mutant s-LNvs projection phenotype. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (magenta) antibodies.

(E) Quantification of horizontal A.U. in Dscam1 mutant flies. Data are presented as mean ± SD. w1118 (n = 22) Dscam121/+ (n = 5), Dscam11/+ (n = 8), Dscam105518/+ (n = 10), Dscam121/ Dscam105518(n = 6), Dscam121/Dscam11 (n = 10), Dscam105518(n = 14). One-way ANOVA with Tukey’s post hoc, ns, p > 0.05, ****p < 0.0001.

(F) Endogenous Dscam1 co-localizes with the s-LNvs axon terminal. Pdf >GFP fly heads were collected at 120 h ALH and stained with anti-Dscam1 (red). The white dotted line indicates the s-LNvs axon.

We further checked s-LNvs axon projection in several well-defined Dscam1 mutants, including Dscam121, Dscam11, and Dscam105518 (Hummel et al., 2003; Schmucker et al., 2000). While all heterozygous mutants displayed normal horizontal axon projection in s-LNvs, trans-heterozygous mutants and homozygotes exhibited shortened horizontal axon projection (Figure 3D-E). Additionally, our immunostaining results indicated concentrated Dscam1 expression at the growth cone of s-LNv axons in 3rd-instar larvae (Figure 3F). Taken together, these data demonstrate that Dscam1 controls the horizontal axon projection of s-LNvs in a cell-autonomous manner.

Dscam1 is a well-recognized cell adhesion molecule that plays a crucial role in axon guidance (Chen et al., 2006; Hummel et al., 2003; Zhan et al., 2004; Zhang et al., 2019). The receptors present on growth cones sensor the extracellular axon guidance molecules to initiate the reorganization of the cellular cytoskeleton, leading to the facilitation of axonal projection. Convincingly, knockdown of the cytoskeletal molecules tsr (the Drosophila homolog of cofilin) (Sudarsanam et al., 2020) and chic (the Drosophila homolog of profilin) (Shields et al., 2014) in s-LNvs recaptured the horizontal axon projection deficits in Dscam1 knockdown flies or Dscam1 mutants (Figure 3-figure supplement 1C-D). Consistently, knockdown of Dock and Pak, the guidance receptor partners of Dscam1 (Schmucker et al., 2000), or SH3PX1, the critical linker between Dscam1 and the cytoskeleton (Worby et al., 2001), phenocopied Dscam1 knockdown flies or Dscam1 mutants (Figure 3-figure supplement 1C-D). Taken together, these findings provide solid evidence to support that Dscam1 signaling as a crucial regulator of s-LNvs axon horizontal projection (Figure 3-figure supplement 1E).

Neuronal-derived Netrins act upstream of Dscam1 to govern the horizontal projection of s-LNvs axon

Two classical guidance molecules, Slit and Netrin, have been shown to specifically bind to the extracellular domain of Dscam1 (Alavi et al., 2016; Andrews et al., 2008). The Drosophila genome contains the sole slit gene and two Netrin genes: Netrin-A (NetA) and Netrin-B (NetB) (Harris et al., 1996; Mitchell et al., 1996). Interestingly, knockdown of both NetA and NetB (hereafter referred to as Netrins), rather than slit, resulted in the defective horizontal axon projection of s-LNvs. Convincingly, NetA,NetB double mutant (Brankatschk & Dickson, 2006), but not NetA or NetB single mutant, exhibited the defective horizontal axon projection of s-LNvs (Figure 4A-B and Figure 4-figure supplement 1). These results imply that two Netrin molecules function redundantly in regulating horizontal axon projection of s-LNvs.

Neuron-derived Netrin guides s-LNvs horizontal projection.

(A) Images of immunostained in w1118, NetAΔ, NetBΔ and NetABΔ fly. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (magenta) antibodies.

(B) Quantification of horizontal A.U. w1118, NetAΔ, NetBΔ and NetABΔ fly. Data are presented as mean ± SD, w1118(n = 17), NetAΔ (n = 15), NetBΔ (n = 24), NetABΔ (n = 23). One-way ANOVA with Dunnett’s post hoc, ns, p > 0.05, ***p < 0.001.

(C) Images of immunostained in nSyb-GAL4 and repo-GAL4 knockdown Netrins. Larvae brains were collected at late 3rd larvae heads were stained with anti-PDF (white) and HRP (magenta) antibodies.

(D) Quantification of horizontal A.U. in nSyb-GAL4 and repo-GAL4 knockdown Netrins fly. Data are presented as mean ± SD. nSyb-GAL4/+ (n = 13), nSyb-GAL4 >NetrinsRNAi (n = 16), repo-GAL4/+ (n = 34), repo-GAL4 >NetrinsRNAi (n = 20). Two-tailed Student’s t tests, ns, p > 0.05, ***p < 0.001.

To determine the source of Netrins, which are known to be secreted axon guidance molecules, we conducted experiments using pan-neuronal (nsyb-GAL4) and pan-glial (repo-GAL4) drivers to selectively knock down Netrins (Figure 4C-D). Strikingly, when Netrins were knocked down in neurons but not in glia, we observed severe defects in the horizontal axon projection of s-LNvs. These findings reveal that neuron-secreted Netrins serve as ligands for Dscam1, controlling the horizontal axon projection of s-LNvs.

Dorsal neuron-secreted Netrins mediate the horizontal axon projection of s-LNvs

To further identify the source of Netrins, we focused on the neurons located in the dorsolateral protocerebrum, where the s-LNvs axons change their projection direction. We first excluded the possibility that MB-secreted Netrins mediate the horizontal axon projection of s-LNvs, as knockdown of Netrins with OK107-GAL4 showed normal horizontal axon projection of s-LNvs (Figure 5A-B).

Netrin secreted by DN neurons guides time-specific horizontal projection of s-LNvs.

(A) Images of immunostained in OK107-GAL4 and per-GAL4, Pdf-GAL80 knockdown Netrins. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (magenta) antibodies.

(B) Quantification of horizontal A.U. in OK107-GAL4 and per-GAL4, Pdf-GAL80 knockdown Netrins fly. Data are presented as mean ± SD. OK107-GAL4/+ (n = 20), OK107-GAL4 >NetrinsRNAi (n = 13). per-GAL4, Pdf-GAL80/+ (n = 18), per-GAL4, Pdf-GAL80 >NetrinsRNAi (n = 16), Two-tailed Student’s t tests for OK107-GAL4 knockdown Netrins, ns, p > 0.05. Mann-Whitney test for per-GAL4, Pdf-GAL80 knockdown Netrins, ****p < 0.0001.

(C) Images of DN ablation in the developing larval stages. Larvae brains were stained with anti-PDF (white) and HRP (magenta) antibodies. Different time indicated hours ALH.

(D) Quantification of horizontal A.U. in Control (per-GAL4, Pdf-GAL80 >GFP) and Ablation (per-GAL4, Pdf-GAL80 >GFP,rpr,hid) flies. Data are presented as mean ± SD. Control: 72 h (n = 5), 96 h (n = 12), Ablation: 72 h (n = 5), 96 h (n = 10). Two-tailed Student’s t tests were used to compare conditions. ns, p > 0.05, ****p < 0.0001.

(E) Images of immunostained in per-GAL4, Pdf-GAL80 overexpress NetB in NetABΔ. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (magenta) antibodies.

(F) Quantification of horizontal A.U. in per-GAL4, Pdf-GAL80 overexpress NetB in NetABΔ. Data are presented as mean ± SD. NetABΔ, UAS-NetB (n = 12), NetABΔ, per-GAL4,Pdf-GAL80 /+ (n = 5), NetABΔ,per-GAL4,Pdf-GAL80 >UAS-NetB (n = 9). One-way ANOVA with Bonferroni post hoc, ns, p > 0.05, ***p < 0.001.

The Dorsal neurons (DNs), a subset of clock neurons, also situate in the dorsolateral protocerebrum. DNs are categorized into three types, DN1, DN2, and DN3 (Reinhard et al., 2022), and previous studies have shown that both DN2 and DN1 form synaptic connections with s-LNvs axon terminals at the adult stage (Schlichting et al., 2022). Interestingly, knockdown of Netrins in a substantial portion of DN2, DN3, and DN1 (Kaneko et al., 1997) (Per-GAL4, Pdf-GAL80;UAS-Netrins-RNAi) resulted in the defective horizontal axon projection of s-LNvs (Figure 5A-B). To further validate the involvement of DNs in mediating the horizontal axon projection of s-LNvs, we conducted cell ablation experiments. Notably, the impairments in the horizontal axonal projection of s-LNvs due to DNs ablation were exclusively observed at 96 hours ALH, which is after the occurrence of horizontal projection (Figure 5C-D). In contrast, ablating crz+ neurons, which occupy a similar location to DNs, had no significant effects on the horizontal axon projection of the s-LNvs (Figure 5-figure supplement 1). Moreover, by expressing NetB in DNs in NetA,NetB double mutants, we successfully restored the defective horizontal axon projection of s-LNvs (NetABΔ;Per-GAL4, Pdf-GAL80;UAS-NetB) (Figure 5E-F). These results demonstrate that DNs secrete Netrins to guide the horizontal axon projection of s-LNvs.

Newborn DNs secret Netrins to regulate the horizontal axon projection of s-LNvs

Finally, we wonder which population of DNs secret Netrins to regulate the horizontal axon projection of s-LNvs. Therefore, we monitored the number and location of DNs during the s-LNvs axon projection and found that the number of DNs significantly increased during this process. Per-GAL4, Pdf-GAL80; UAS-mCD8::GFP only labeled 4-5 cells at 48 hours ALH, and raise to 10-15 by 72 hours, and subsequently increased to approximately 25 by 96 hours ALH (Figure 6A-B). It is worth noting that the location of these newly formed DNs resides lateral to the transition point of axon projection direction of the s-LNvs, while maintaining a basic parallelism with the horizontal axon projection of the s-LNvs (Figure 6A). The sharp increase in the number of DNs coincides remarkably with the timing of the switch in axon projection direction of the s-LNvs.

Dynamic changes in DN neurons coordinated with s-LNvs axonal targeting.

(A) Top: Images of s-LNvs axon and DNs growth process in the developing larval stages. Larvae brains were stained with anti-PDF (white) and GFP (green) antibodies. Different time indicated hours ALH are shown. Bottom: schematic diagram of s-LNvs horizontal projection and the corresponding increase in the number of DN neurons. s-LNvs axon (gray), DN neurons (green).

(B) Quantification of the number of DN neurons labeled by per-GAL4, Pdf-GAL80 at different developmental stages. Data are presented as mean ± SD.

(C) Images of newborn DN neurons were colocalized with Netrin-B at different developmental times. Larvae brains were stained with anti-GFP (green), mcherry (red) and PDF (white) antibodies. Different time indicated hours ALH. The white arrow demarcates the co-localization of red and green signals.

(D) Schematic representation s-LNvs projection directional transition and the corresponding increase in the number of DN neurons. s-LNvs axon (gray), DN neurons (green), newborn DN neurons (light green at 72 h, orange at 96 h).

Next, we asked whether these newly generated DNs expressed Netrins. To visualize the expression of endogenous NetB, we applied the fly strains that insert either GFP or myc tag in NetB gene. At 72 hours ALH, we were unable to detect any NetB signals within the DNs. In contrast, at 96 hours ALH, we easily detected prominent NetB signals in approximately 6-8 newborn DNs (Figure 6C-D, Figure 6-figure supplement 1A).

Previous studies have shown that DSCAM is involved in Netrin-1-mediated axonal attraction (Andrews et al., 2008; G. Liu et al., 2009; Ly et al., 2008). In addition, DSCAM also functions as a repulsive receptor, associating with Uncoordinated-5C (UNC5C) to mediate Netrin-1-induced axon growth cone collapse (Purohit et al., 2012). To dissect the role of Netrin signaling, we ectopically expressed Netrins in neurons marked by R78G02-GAL4 (Jenett et al., 2012; Suzuki et al., 2022), located in front of the horizontal projection of s-LNv axons, closer to the midline (Figure 6-figure supplement 1B). Expression of either NetA or NetB in R78G02-GAL4-labeled neurons significantly suppressed the horizontal axon projection of s-LNvs (Figure 6-figure supplement 1B-C). Taken together, these findings reveal that newborn DNs secrete Netrins to orchestrate the transition of axon projection in s-LNvs from a vertical to a horizontal direction (Figure 7).

Dynamic cellular molecular environment during the s-LNv projection directional shift.

Cartoon depicting the dynamic cellular molecular microenvironment during the axonal projection directional shift from vertical to horizontal projection in s-LNvs. 72 h and 96 h represent 72 h and 96 h after larval hatching, respectively. s-LNvs (blue), neuropile (pink), brain lobe (gray circule), DNs (orange and light orange circles), optic lobe (gray dotted line), Netrin (Orange combination molding), Dscam1 (pruple).

Discussion

The mechanisms underlying axonal responses to intricate microenvironments and the precise development of axons within the brain have long remained enigmatic. In the past, researchers have identified a large number of axon guidance cues and receptors using different models in vivo and in vitro (Esther T Stoeckli, 2018). In addition, transient cell-cell interactions through intermediate targets play a crucial role in guiding axonal projection step by step towards its final destination (Chao et al., 2009; Garel & Rubenstein, 2004). The landmark midline, a crucial intermediate target, serves as a model in the majority of early studies exploring the directional transitions of axonal projections (Evans & Bashaw, 2010). However, so many neural projections do not cross the midline that it is challenging to understand how these axon projections are guided (Goodhill, 2016). In our study, we established an excellent model to investigate the mechanism of axonal projection direction switch by the Drosophila s-LNvs. We discovered a coordinated growth pattern between the vertical projection of s-LNvs and the MB calyx. Furthermore, the synergistic interaction between Dscam1 expressed in s-LNvs and the emerging DN-secreted Netrins precisely modulates the transition of s-LNv axons from a vertical to a horizontal projection within a specific time window (Figure 7). These findings reveal the mechanism behind the transition of axonal projection direction, emphasizing the significance of developmental microenvironments in ensuring precise axon projection.

The dependence of s-LNv vertical projection on mushroom body calyx expansion

During our monitoring of the axonal projection of s-LNvs, we observed that the axon terminals reached the dorsolateral brain area at an early stage but continued their vertical growth. This phenomenon raises the question of what drives this growth. The mushroom body, a sophisticated central hub within the fruit fly’s brain, exhibits proximity to the projection of s-LNv axons in spatial arrangement. Upon careful examination, we found a positive correlation between the vertical length increase of s-LNv axons and the growth of the mushroom body calyx (Figure 2). Remarkably, when we selectively removed KC cells, we observed a striking effect on the axonal projection of s-LNvs. The s-LNv axonal projections either stalled at the initial stage or even completely disappeared (Figure 2 and data not shown). Hence, the vertical projection of s-LNv axons is dependent on mushroom body calyx expansion. Our hypothesis is supported by the findings of Helfrich-Förster et al., who reported locomotor activity and circadian rhythm defects in some MB mutants (HELFRICH-FÖRSTER et al., 2002), potentially attributed to the abnormality of s-LNvs axon. Moreover, it was observed that the MB mutants had minimal impact on the l-LNvs. This can be attributed to the fact that l-LNvs fibers emerge at a later stage during pupal development and are situated at a spatial distance from the mushroom body.

Unfortunately, we did not identify any molecules that have a discernible impact on the vertical projection of s-LNv axons through screening. This indicates that the successful completion of the vertical projection process may play a pivotal role in determining the overall existence of s-LNv axonal projections. Furthermore, the dorsal projection of s-LNv axons intersects with the ventral projection of neurons such as DN1 in the dorsolateral region adjacent to the calyx (Keene et al., 2011), resembling the corpus callosum in mammals (Fothergill et al., 2014; Hutchins et al., 2011; Keeble et al., 2006; Piper et al., 2009; Unni et al., 2012). Thus, during early developmental stages, multiple guiding cues may redundantly function between the mushroom body calyx and neighboring neural processes to ensure the smooth progression of neural development and establish a stable brain structure. However, to gain a more comprehensive understanding, it is worthwhile to explore whether the development of the mushroom body calyx influences the projection of other neurons traversing the same territory and the guidance cues involved. Insights from existing single-cell sequencing data obtained from multiple stages of larval life (Avalos et al., 2019; Corrales et al., 2022), might offer some clues and indications.

Netrin/Dscam signaling specifically controls the horizontal projection of s-LNv axons

In this paper, when we specifically knockdown Dscam1 in s-LNvs, we observed a significant defect in the axonal horizontal projection. This suggests that Dscam1 autonomously regulates the horizontal axonal projection of s-LNvs (Figure 3). Our findings align with a recent study that reported abnormal horizontal axonal projection of s-LNv neurons in adult flies when Dscam1 mRNAs lack the long 3’ UTR (Zhang et al., 2019).

The extracellular domain of Dscam1 has been verified to be capable of binding to Netrin or Slit (Alavi et al., 2016). In Drosophila, at the midline of the embryonic central nervous system, Dscam1 forms a complex with Robo1 to receive the Slit signal and promote the growth of longitudinal axons (Alavi et al., 2016). In adult flies, Slit/Robo signaling restricts the medial growth of s-LNv axons (Oliva et al., 2016). However, we found that reducing Netrins levels, rather than slit, led to a decrease in the horizontal projection distance of s-LNv axons. Moreover, in the NetABΔ double mutant, the extent of defects in the horizontal axonal projection of s-LNv neurons is similar to that observed when knocking down Dscam1 in s-LNvs. Indeed, previous studies have provided evidence showcasing the widespread involvement of Dscam1 as a receptor for Netrin in mediating axon growth and pathfinding (G. Liu et al., 2009; Ly et al., 2008; Matthews & Grueber, 2011). When we disturbed the expression of Netrin signals in the axon targeting microenvironment, it was also sufficient to cause abnormal s-LNv projection (Figure 6-figure supplement 1B-C). Therefore, our findings suggest that Netrin, as a ligand for Dscam1, regulates the process of switching axon projection direction (Figure 4).

The newborn DNs-secreted Netrin coordinates with s-LNv-expressed Dscam1 to switch the projection direction of s-LNv axons

While the midline serves as a prominent landmark, it presents a daunting challenge to comprehend the guidance and directional transitions of axons in many neural projections that do not actually cross this central axis (Evans & Bashaw, 2010). Following vertical projection, the s-LNvs growth cones remain within the dorsolateral area for at least 48 hours before initiating horizontal projection (Figure 1). This phenomenon aligns harmoniously with the outcomes observed in earlier studies that the axonal growth of other types of neurons slows down at the specific choice point, the midline (Bak & Fraser, 2003; Godement et al., 1994; T. Li et al., 2021), indicating that the dorsolateral area serves as an intermediate targets to facilitate the subsequent phase of s-LNvs projection journey.

Extracellular cues released by the final or intermediate targets play a crucial role in guiding axonal projection. The axon terminals of s-LNvs showed close spatial associations with the somas and processes of DNs. In this study, the knockdown of Netrins specifically in per+,Pdf- neurons, but not in the mushroom body, resulted in axon projection defects in s-LNvs (Figure 5). Furthermore, the ablation of DNs resulted in the inhibition of horizontal growth of s-LNvs axons (Figure 5). These findings suggest that the Netrin signaling microenvironment secreted by DNs is involved in regulating the horizontal projection of s-LNv axons.

Ideas and Speculation

The emergence of NetB-positive DNs and the aberrant axonal projection of s-LNv neurons caused by DN ablation occur concurrently during development (Figure 6). In Drosophila, three types of circadian oscillator neurons, including two DN1s, two DN2s, and 4-5 s-LNvs, can be identified as early as embryonic stage 16 (Houl et al., 2008). The DN3 group, which represents the largest contingent within the central circadian neuron network with over 35 neurons, emerges during the larval stage (T. Liu et al., 2015). Little is known about the function of DN3 neurons, particularly during the larval stage. Due to the lack of cell-specific labeling tools, we can only speculate about the identity of these newly generated Netrin-secreting neurons as DN3s based on their spatial distribution. Taken together, these results unveil a novel regulatory mechanism where axons, during the process of pathfinding, await guidance cues from newly born guidepost cells. This enables a switch in the direction of axon projection, facilitating the axon’s subsequent journey.

Supporting information

eLife Assessment

This study provides valuable insights into the mechanism of axonal directional changes, utilizing sLNv neurons as a model. The data were collected and analysed using solid methodology, although the conceptual framing of the work and contextualization of the results require revision and reassessment. The study holds potential interest for neurobiologists focusing on axonal growth and development.

Reviewer #1 (Public Review):

Summary:

The mechanisms of how axonal projections find their correct target requires the interplay of signalling pathways, and cell adhesion that act over short and long distances. The current study aims to use the small ventral lateral clock neurons (s-LNvs) of the Drosophila clock circuit as a model to study axon projections. These neurons are born during embryonic stages and are part of the core of the clock circuit in the larval brain. Moreover, these neurons are maintained through metamorphosis and become part of the adult clock circuit. The authors use the axon length by means of anti-Pdf antibody or Pdf>GFP as a read-out for the axonal length. Using ablation of the MB- the overall target region of the s-LNvs, the authors find defects in the projections. Next, by using Dscam mutants or knock-down they observe defects in the projections. Manipulations by the DNs - another group of clock neurons- can induce defects in the s-LNvs axonal form, suggesting an active role of these neurons in the morphology of the s-LNvs.

Strengths:

The use of Drosophila genetics and a specific neural type allows targeted manipulations with high precision.

Proposing a new model for a small group of neurons for axonal projections allows us to explore the mechanism with high precision.

Weaknesses:

It is unclear how far the proposed model can be seen as developmental.

The study of changes in fully differentiated and functioning neurons may affect the interpretation of the findings.

Reviewer #2 (Public Review):

Summary:

The paper from Li et al shows a mechanism by which axons can change direction during development. They use the sLNv neurons as a model. They find that the appearance of a new group of neurons (DNs) during post-embryonic proliferation secretes netrins and repels horizontally towards the midline, the axonal tip of the LNvs.

Strengths:

The experiments are well done and the results are conclusive.

Weaknesses:

The novelty of the study is overstated, and the background is understated. Both things need to be revised.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The mechanisms of how axonal projections find their correct target requires the interplay of signalling pathways, and cell adhesion that act over short and long distances. The current study aims to use the small ventral lateral clock neurons (s-LNvs) of the Drosophila clock circuit as a model to study axon projections. These neurons are born during embryonic stages and are part of the core of the clock circuit in the larval brain. Moreover, these neurons are maintained through metamorphosis and become part of the adult clock circuit. The authors use the axon length by means of anti-Pdf antibody or Pdf>GFP as a read-out for the axonal length. Using ablation of the MB- the overall target region of the s-LNvs, the authors find defects in the projections. Next, by using Dscam mutants or knock-down they observe defects in the projections. Manipulations by the DNs - another group of clock neurons- can induce defects in the s-LNvs axonal form, suggesting an active role of these neurons in the morphology of the s-LNvs.

Strengths:

The use of Drosophila genetics and a specific neural type allows targeted manipulations with high precision.

Proposing a new model for a small group of neurons for axonal projections allows us to explore the mechanism with high precision.

Weaknesses:

It is unclear how far the proposed model can be seen as developmental.

The study of changes in fully differentiated and functioning neurons may affect the interpretation of the findings.

We appreciate the reviewer's feedback on the strengths and weaknesses of our study.

We acknowledge the strengths of our research, particularly the precision afforded by using Drosophila genetics and a specific neural type for targeted manipulations, as well as the proposal of a new model for studying axonal projections in a small group of neurons.

We understand the concerns about the developmental aspects of our proposed model and the use of Pdf-GAL4 >GFP as a read-out for the axonal length (revised manuscript Figure 1--figure supplement 1). However, even with the use of Clk856-GAL4 that began to be expressed at the embryonic stage (revised manuscript Figure 3--figure supplement 1) to suppress Dscam expression, the initial segment of the dorsal projection of s-LNvs (the vertical part) remained unaffected. Instead, the projection distance is severely shortened towards the midline, and this defect persists until the adult stage. It is for this reason that we delineate the dorsal projections of s-LNvs into two distinct phases: the vertical and horizontal parts, rather than a mere expansion in correspondence with the development of the larval brain.

Thank you for your valuable feedback, and we have incorporated these considerations into our revised manuscript to enhance the clarity and depth of our research.

Reviewer #2 (Public Review):

Summary:

The paper from Li et al shows a mechanism by which axons can change direction during development. They use the sLNv neurons as a model. They find that the appearance of a new group of neurons (DNs) during post-embryonic proliferation secretes netrins and repels horizontally towards the midline, the axonal tip of the LNvs.

Strengths:

The experiments are well done and the results are conclusive.

Weaknesses:

The novelty of the study is overstated, and the background is understated. Both things need to be revised.

We appreciate your acknowledgment that the experiments were well-executed and the results conclusive. This validation reinforces the robustness of our findings.

We take note of your feedback regarding the novelty of the study being overstated and the background being understated. While axonal projections navigate without distinct landmarks, like the midline or the layers, columns, and segments, they pose more challenges and uncertainties. As highlighted, our key contribution lies in elucidating how axonal projections without clear landmarks are guided, with our research demonstrating how a newly formed cluster of cells at a specific time and location provides the necessary guidance cues for axons.

We value your insights, and we have carefully addressed these points in our manuscript revision to improve the overall quality and presentation of our research.

Recommendations For The Authors:

Reviewer #1 (Recommendations For The Authors):

The overall idea of using the s-LNvs as a model is indeed intriguing. There are genetic tools available to tackle these cells with great precision.

However, based on the stage at which these cells are investigated raises some issues, that I feel are critical to be addressed.

These neurons develop their axonal projections during embryogenesis and are fully functioning when the larvae hatch, thus to investigate axonal pathfinding one would have to address embryonic development.

The larval brain indeed continues to grow during larval life, however extensive work from the Hartenstein lab, Truman lab, and others have shown that the secondary (larval born) neurons do not yet wire into the brain, but stall their axonal projections.

It is thus quite unclear, what the authors are actually studying.

One interpretation could be that the authors observe changes in axon length due to morphological changes in the brain. Indeed, the fact that the MB expands the anatomy of the surrounding neuropil changes too.

Moreover, it is unclear when exactly the Pdf-Gal4 (and other drivers) are active, thus how far (embryonic) development of s-LNvs is affected, or if it's all happening in the differentiated, functioning neuron. (Gal4 temporal delay and dynamics during embryonic development may further complicate the issue). As far as I am aware the MB drivers might already be active during embryonic stages.

Since the raised issue is quite fundamental, I am not sure what might be the best and most productive fashion to address this.

Eg. either to completely re-focus the topic on "neural morphology maintenance" or to study the actual development of these cells.

We thank the reviewer for the detailed and insightful feedback on our study. We have tested whether Pdf-Gal4 could effectively label s-LNv, and tracked the s-LNv projection in the early stage after larvae hatching. We did not observe the PDF antibody staining signal and the GFP signal driven by Pdf-GAL4 when the larvae were newly hatched. At 2-4 hours ALH, PDF signals were primarily concentrated at the end of axons, while GFP signals were mainly concentrated at the cell body. Helfrich-Förster initially detected immunoreactivity for PDF in the brains approximately 4-5 hours ALH. The GFP signal expressed by Pdf-GAL4 driver does have signal delay. However, at 8 hours ALH, the GFP signal strongly co-localized with the PDF signal within the axons (see revised manuscript lines 98-101) (Figure 1—figure supplement 1).

Based on previous research findings and our staining of Clk856-GAL4 >GFP, it is indeed confirmed that the dorsal projection of s-LNvs in Drosophila is formed during the embryonic stage (Figure 3—figure supplement 1). The s-LNvs in first-instar larval Drosophila are capable of detecting signal output and may play a role in regulating certain behaviors. Our selection of tools for characterizing the projection pattern of s-LNv was not optimal, leading us to overlook the crucial detail that the projection had already formed during its embryonic stage.

However, even when employing Clk856-GAL4 to suppress Dscam expression from the embryonic stage, the initial segment of the dorsal projection of s-LNvs (the vertical part) remains unaffected. Instead, the projection distance is severely shortened towards the midline, and this defect persists until the adult stage. It is for this reason that we delineate the dorsal projections of s-LNvs into two distinct phases: the vertical and horizontal parts, rather than a mere expansion in correspondence with the development of the larval brain.

From the results searched in the Virtual Fly Brain (VFB) database (https://www.virtualflybrain.org/), it is clear that the neurons that form synaptic connections with s-LNvs at the adult stage are essentially completely different from the neurons that are associated with them at the L1 larval stage. Thus, most neurons that form synapses with s-LNvs in the early larvae either cease to exist after metamorphosis or assume other roles in the adult stage. Similar to the scenario where Cajal-Retzius cells and GABAergic interneurons establish transient synaptic connections with entorhinal axons and commissural axons, respectively, these cells form a transient circuit with presynaptic targets and subsequently undergo cell death during development. In our model, the neurons that synapse with s-LNvs in early development serve as "placeholders," offering positive or negative cues to guide the axonal targeting of s-LNvs towards their ultimate destination.

Thank you again for your valuable feedback, and we have incorporated these considerations into our revised manuscript to enhance the clarity and depth of our research.

Reviewer #2 (Recommendations For The Authors):

Major:

In the introduction too many revisions are cited and very few actual research papers. This should be corrected and the most significant papers in the field should be cited. For example, there is no reference to the pioneering work from the Christine Holt lab or the first paper looking at axon guidance and guideposts by Klose and Bentley, Isbister et al 1999.

The introduction should encapsulate the actual knowledge based on actual research papers.

We acknowledge your concern regarding the citation of review papers rather than primary research papers in the introduction. Following your suggestion, we have revised the introduction section to incorporate references to relevant research papers.

In the introduction and discussion: The authors cite revisions where the signals that guide axons across different regions including turning are shown and they end up saying: "However, how the axons change their projection direction without well-defined landmarks is still unclear." I think the sentence should be changed. Many things are still not clear but this is not a good phrasing. Maybe they could focus on their temporal finding?

We appreciate the reviewer's feedback and insightful suggestions. We agree that emphasizing the temporal aspect is crucial in our study. However, we also recognize the significance of understanding the origin of signals that guide axonal reorientation at specific locations. While axonal projections navigating without distinct landmarks pose more challenges and uncertainties compared to those guided by prominent landmarks like the midline, our research demonstrates the crucial role of a specific cell population near turning points in providing accurate guidance cues to ensure precise axonal reorientation. We have revised our phrasing in the introduction and discussion to better reflect these key points (see revised manuscript lines 69-71 and 350-354). Thank you for highlighting the significance of focusing on our temporal findings and the complexities involved in studying axonal projection.

Many rather old papers have looked into the effect of repulsive guideposts to guide axon projections. In particular, I can think of the paper from Isbister et al. 1999 (DOI: 10.1242/dev.126.9.2007) that not only shows how semaphoring guides Ti axon projection but also shows how the pattern of expression of sema 2a changes during development to guide the correct projection. I really think that the novelty of the paper should be revised in light of the actual knowledge in the field.

We appreciate the reviewer's reference to the seminal work by Isbister et al. (1999) and the importance of guidepost cells in axon projection guidance, which we have already cited in our revised manuscript. It is crucial to recognize that segmented patterns such as the limb segment traversed by Ti1 neuron projections or neural circuits formed in a layer- or column-specific manner also serve as intrinsic "guideposts," offering valuable insights into axonal pathfinding processes. In our model, explicit guidance cues are lacking. As highlighted, our key contribution lies in elucidating how axonal projections without clear landmarks are guided, with our research demonstrating how a newly formed cluster of cells at a specific time and location provides the necessary guidance cues for axons (see revised manuscript lines 350-354). We have ensured that our revised manuscript reflects these insights and emphasizes the significance of studying axonal guidance in the absence of distinct guideposts. Thank you for underscoring these essential points, which enhance our understanding of axonal projection dynamics.

Minors:

Line 54, the authors start talking about floorplate at the end of a section on Drosophila. Please use “In vertebrates”, or “in invertebrates” or “in Drosophila” etc.. when needed to put things in context.

We thank the reviewer for this suggestion and have modified this sentence. Please refer to lines 62-63 of the revised manuscript.

Line 69: many factors change the axonal outgrowth. The authors are missing the paper from Fernandez et al. 2020, who have shown that unc5 the receptor of netrin induces the stalling for sLNvs projections before the turn. https://doi.org/10.1016/j.cub.2020.04.025

We thank the reviewer for this suggestion and have added this research article. Please refer to line 79 of the revised manuscript.

Line 99: "precisely at the pivotal juncture". It I hard to see how it was done in the figures shown. Can the authors add a small panel with neuronal staining showing this (please no HRP)?

For all figures, tee magenta is too strong and it is really hard to see the sLNvs projections. Can this be sorted, please?

We have depicted the pivotal juncture in the schematic diagram on the left side of Figure 1C. Additionally, we have included a separate column of images without HRP in Figure 1A. Moreover, we have modified the pseudo-color of HRP from magenta to blue to enhance the visualization of the s-LNv projection. The figure legends have also been correspondingly modified.

Line 407: Spatial position relationship between calyx and s-LNvs. OK107-GAL4 labels ... calyx and s-LNvs labeled by, which which.

We have modified it according to your suggestion. Please refer to lines 430-432 of the revised manuscript.

Line 137 typo RPRC

We thank the reviewer for noticing this mistake, which has now been corrected. Please refer to line 148-149 of the revised manuscript.

Section 158-164. the paper from Zhang et al 2019 needs to be cited since they have found the same effect of decreasing Dscam even if they didn't think about horizontal projection.

Thanks to the suggestion, we have included in the manuscript the phenotype observed by Zhang et al. (2019) upon knocking down Dscam1-L in adults. Please refer to lines 170-172 of the revised manuscript.

Line 176: typo senses (instead of sensor).

Thank you for pointing out our mistake. We have modified it according to your suggestion. Please refer to line 189 of the revised manuscript.

Line 193: more than Interesting it is Notable. Add "ubiquitus" knockdown.

Thank you for the suggestion. We have included the word "ubiquitus" to enhance the precision of the narrative. Please refer to line 206 of the revised manuscript.

Line 224: the pattern of expression of the crz cells is not visible where the projections of sLNvs are located. Are they in that region? Or further away?

We've changed the pseudo-color of HRP, and in the updated Figure 5- figure supplement 1, you can see the projection pattern of crz+ cells, positioned close to the end of the s-LNv axon terminal.

Line 243: applied? Do you mean "used"

Thank you for the suggestion. We have revised it at line 256.

Figure 5 Sup1: the schematic shows DNs proliferation that is not visible on the GFP image. Please comment.

We have modified the Figure 5 figure supplementary 1 for 120 h per-GAL4, Pdf-GAL80 >GFP expression pattern. Due to the strong GFP intensity in some DN neurons, there was a loss of GFP signal. Additionally, in Figure 6 figure supplementary 1, we have added co-localization images of DN and s-LNv at 72 h and 96 h. To better illustrate the co-localization information, we have shown only a portion of the layers in the right panel. We hope these additions clarify your concerns.

Line 251: cite Fernandez et al. 2020 with Purohit et al 2012.

We have modified it according to your suggestion. Please refer to line 264 of the revised manuscript.

Line 272: you have not shown synergistic effects because you have not modulated both pathways at the same time. You should talk about complementary.

We have modified it according to your suggestion at lines 25, 285, 439.

Version 2

Abstract

Abstract

Axon projection is a spatial and temporal-specific process in which the growth cone receives environmental signals guiding axons to their final destination. However, the mechanisms underlying changes in axonal projection direction without well-defined landmarks remain elusive. Here, we present evidence showcasing the dynamic nature of axonal projections in Drosophila’s small ventral lateral clock neurons (s-LNvs). Our findings reveal that these axons undergo an initial vertical projection in the early larval stage, followed by a subsequent transition to a horizontal projection in the early-to-mid third instar larvae. The vertical projection of s-LNv axons correlates with mushroom body calyx expansion, while the s-LNv-expressed Down syndrome cell adhesion molecule (Dscam1) interacts with Netrins to regulate the horizontal projection. During a specific temporal window, locally newborn dorsal clock neurons (DNs) secrete Netrins, facilitating the transition of axonal projection direction in s-LNvs. Our study establishes a compelling in vivo model to probe the mechanisms of axonal projection direction switching in the absence of clear landmarks. These findings underscore the significance of dynamic local microenvironments in the complementary regulation of axonal projection direction transitions.

Introduction

During nervous system development, neurons extend axons to reach their targets in order to build functional neural circuits (Bentley & Caudy, 1983; Klose & Bentley, 1989; Mann et al., 2002; Shi et al., 2023). The growth cone, which is specialized at the tip of the extending axon, is critical for receiving multiple guidance signals from the external environment to guide axon projection (Agi et al., 2024; Lowery & Vactor, 2009). Guidance cues regulate cytoskeletal dynamics through growth cone-specific receptors, steering axons via attractive or repulsive signals (Tessier-Lavigne & Goodman, 1996, Araújo & Tear, 2003; Koch et al., 2012; Kolodkin & Tessier-Lavigne, 2011; Esther T Stoeckli, 2018; Zang et al., 2021). Over the last 30 years, several guidance cues have been identified and divided into four classical families: Semaphorins (A L Kolodkin et al., 1992; Luo et al., 1993; Pasterkamp, 2012; Raper, 2000; Winberg et al., 1998), Ephrins (Cheng et al., 1995; Kania & Klein, 2016; SPERRY, 1963; Wilkinson, 2001), Netrins (Colamarino & Tessier-Lavigne, 1995; Hong et al., 1999; Kennedy et al., 1994; Meijers et al., 2020; Pan et al., 2023; Serafini et al., 1994), and Slits (Agi et al., 2024; Battye et al., 1999; Kidd et al., 1999; Seeger et al., 1993; Tessier-Lavigne & Goodman, 1996). The emergence of other kinds of guidance cues, including morphogens (Charron et al., 2003; Ciani et al., 2004; Colavita et al., 1998; Wen et al., 2007; Yoshikawa et al., 2003), growth factors (Barde et al., 1982; Berkemeier et al., 1991; Leibrock et al., 1989; Maisonpierre et al., 1990; Short et al., 2021), and cell adhesion molecules (Rader et al., 1996; RADER et al., 1993; Williams et al., 1994; Wong et al., 1995), increases the complexity of axon pathfinding.

When going through a long-distance pathfinding process in the complex and highly dynamic in vivo environment, axons frequently undergo multiple changes in their projection directions (Chitnis & Kuwada, 1991; Hidalgo & Brand, 1997; Hutter, 2003; Isbister et al., 1999; McConnell et al., 1989). Crucially, intermediate targets play a pivotal role by providing vital guiding information that enables axons to transition their projection directions and embark upon the subsequent stages of their journey, ultimately leading them to their final destination (Bartoe et al., 2006; de Ramon Francàs et al., 2017; Esther T Stoeckli, 2018; Timofeev et al., 2012, 2012). Certain intermediate targets can respond to the growth cone’s signaling to maintain a dynamic balance between attractive and repulsive forces (Bron et al., 2007; Martins et al., 2022). One of the well-studied intermediate target projection types is the midline cross model (Bernhardt et al., 1992; Evans et al., 2015; Holley, 1982). As a clear landmark, axons need to project to the midline first and then leave it immediately (Bovolenta & Dodd, 1990; Moreno-Bravo et al., 2019; Pignata et al., 2016, Tulloch et al., 2019). In vertebrates, the midline floor plate cells secrete long-range attractive signals to attract commissural axons towards them (Kennedy et al., 1994; Serafini et al., 1996). Once there, these cells emit signals of rejection to prevent excessive closeness of axons, thus aiding the axon in departing the floor plate to reach its intended destination (Kidd et al., 1998; Long et al., 2004; E T Stoeckli et al., 1997). Comparatively, the typical axonal projection process within the in vivo milieu undergoes multiple transitions in projection direction without a distinct landmark, such as the midline (Garel & López-Bendito, 2014; Gezelius & López-Bendito, 2017; López-Bendito & Molnár, 2003; Molnár et al., 2012). Nevertheless, the precise mechanisms underlying the alteration of axonal projection direction in the absence of distinct landmarks remain to be fully elucidated.

Drosophila small ventral lateral clock neurons (s-LNvs) exhibit the typical tangential projection pattern in the central brain (Agrawal & Hardin, 2016; Gummadova et al., 2009; Hardin, 2017; Helfrich-Förster, 1997). The s-LNv axons originate from the ventrolateral soma and project to the dorsolateral area of the brain, and subsequently undergo a axon projection direction shift, projecting horizontally toward the midline with a short dorsal extension (Helfrich-Förster et al., 2007). Several studies have reported the impact of various factors, including the Slit-Robo signal (Oliva et al., 2016), Unc5 (Fernandez et al., 2020), Lar (leukocyte-antigen-related) (Agrawal & Hardin, 2016), dfmr1 (Okray et al., 2015), dTip60 (Pirooznia et al., 2012), and Dscam1 long 3’ UTR (Zhang et al., 2019), on the axon outgrowth of s-LNvs in the adult flies’ dorsal projection. However, the mechanism underlying the projection direction switch of s-LNv axons remains largely unknown. Here, we show that s-LNvs and newborn dorsal clock neurons (DNs) generate spatiotemporal-specific guidance cues to precisely regulate the transition of projection direction in s-LNv axons. This regulation uncovers unexpected interdependencies between axons and local microenvironment dynamics during the navigation process, enabling accurate control over axon projection.

Results

s-LNvs axons change their projection direction in the early-to-mid third instar larvae

The stereotypical trajectory of the s-LNvs axon projection can be succinctly characterized as an initial vertical extension originating from the ventrolateral brain, followed by a directional pivot at the dorsolateral protocerebrum, ultimately leading to a horizontal projection towards the midline. Immunoreactivity for pigment-dispersing factor (PDF) (Cyran et al., 2005), representing the pattern of LNv neurons, is initially detected in the brains of 1st-instar larvae 4-5 hours after larval hatching (ALH) (Helfrich-Förster, 1997). To elucidate the intricate axon pathfinding process during development, we employed a dual-copy Pdf-GAL4 to drive the expression of double-copy UAS-mCD8::GFP, visualizing the s-LNvs at the larval stage (Figure 1A). We first tested whether Pdf-Gal4 could effectively label s-LNv, and tracked the s-LNv projection in early larvae after embryo hatching. At 8 hours ALH, the GFP signal strongly co-localized with the PDF signal within the axons (Figure 1-figure supplement 1). This remarkable co-localization continued throughout the larval stage, particularly concentrated at the axon’s terminal end (Figure 1A).

s-LNvs axon projection dynamics during the larval stage.

(A) Images showing the growth process of s-LNvs during larval development. Larval brains were stained with anti-PDF (white) and HRP (magenta) antibodies. Different time points indicate the hours after larval hatching (ALH).

(B) Left panel: schematic diagram illustrating the method used to measure the degree of vertical projection. One hemisphere of the larvae brain is depicted. larval neuropil (gray), s-LNvs (green), optic lobe (blue). Right panel: graphs showing the average vertical projection length at different developmental stages, presented as mean ± SD.

(C) Left panel: schematic diagram illustrating the method used to measure the degree of horizontal projection. Right panel: graphs showing the average horizontal projection (A.U.: arbitrary unit) at different developmental stages, presented as mean ± SD. The red line segment indicates the stage at which the axonal projection undergoes a directional transition.

(D) Schematic representation of s-LNvs vertical to horizontal projection directional shift during 72-96 hours ALH. One hemisphere of the larvae brain (gray), larval neuropil (white), s-LNvs (green).

For (B), (C), 8 h (n = 7), 24 h (n = 10), 48 h (n = 14), 72 h (n = 16), 96 h (n = 13), 120 h (n = 16).

We then quantified the process of s-LNvs axonal projection. The vertical projection distance of the s-LNvs axon was ascertained by measuring the vertical span between the lowest point of the optical lobe and the highest point of the axon’s projection terminal. Statistical analyses revealed a steadfast linear increase in the vertical projection distance of the s-LNvs axon as larval development progressed (Figure 1A-B). To determine the horizontal projection distance of the s-LNvs axon, we applied the methodology employed by Olive et al. for measuring adult flies (Oliva et al., 2016). A tangent line is drawn precisely at the pivotal juncture where the s-LNvs axon undergoes a shift in its projection direction. We define the Arbitrary Unit (A.U.) as the ratio between the horizontal distances from this point to both the end of the axon projection and the midline (Figure 1A and 1C). Remarkably, at the time points of 8, 24, 48, and 72 hours ALH, the A.U. value for s-LNvs axon projection in the horizontal direction remained relatively stable at approximately 0.13. However, at 96 and 120 hours ALH, a significant and striking increase in the A.U. value ensued, eventually reaching approximately 0.25 (Figure 1C). These findings suggest that the precise transition of s-LNv axon projection from vertical towards the horizontal direction occurs in the early-to-mid third instar larvae, specifically during 72 to 96 hours ALH (Figure 1D).

Vertical projection of s-LNvs axons correlate with mushroom body calyx expansion

The axons of s-LNv project dorsally, coming in close proximity to the dendritic tree of the mushroom body (MB), specifically the calyx (Figures 2A and 2C). This spatial arrangement facilitates potential interactions between the s-LNvs axon and the MB. MB predominantly comprises 2500 intrinsic neurons, known as Kenyon Cells (KCs) (Puñal et al., 2021). The development of the MB is a dynamic process characterized by three primary types of KCs: α/β, α’/β’, and γ, each following distinct temporal schedules of birth. The γ cells emerge from the embryonic stage until early 3rd-instar larval phase, while the α’/β’ cells are born during the latter half of larval life, and the α/β cells appear in the post-larval stage (Lee et al., 1999; Puñal et al., 2021). We noticed that s-LNv axons exhibit only vertical growth before early 3rd-instar larval stage, which coincides with the emergence of γ KCs. Consequently, we have contemplated the potential correlation between the vertical projection of s-LNv axons and the growth of the mushroom body.

Development of vertical projection process alongside Mushroom Body growth.

(A) Spatial position relationship between calyx and s-LNvs. Top: OK107-GAL4 labels the calyx of MB γ and α’β’ neurons, while s-LNvs are labeled by the PDF antibody. Bottom: Tab2-201Y-GAL4 labels calyx of MB γ neurons, with s-LNvs labeled by the PDF antibody. Larvae brains were stained with anti-PDF (green) and GFP (magenta) antibodies. Different time indicated hours ALH. The white dotted line indicates the calyx region.

(B) Graphs showing the mushroom body calyx’s area and vertical projection length averaged over different development stage are presented as mean ± SD. OK107: 24 h (n = 10), 48 h (n = 8), 72 h (n = 12), 96 h (n = 12), 120 h (n = 9). Pearson’s r = 0.9987, P < 0.0001. Tab2-201Y: 24 h (n = 6), 48 h (n = 12), 72 h (n = 13), 96 h (n = 13), 120 h (n = 11). Pearson’s r = 0.9963, P = 0.0003.

(C) Schematic representation of complementary development of s-LNvs vertical projection and mushroom body calyx. Mushroom body (purple), s-LNvs (Green).

(D) Images of mushroom body ablation in the developing larval stages. Larvae brains were stained with anti-PDF (white) and HRP (blue) antibodies. Different time indicated hours ALH.

(E) Quantification of vertical projection length in Control (Tab2-201Y >GFP) and Ablation (Tab2-201Y >GFP,rpr,hid) flies. Data are presented as mean ± SD. Control: 24 h (n = 6), 48 h (n = 14), 72 h (n = 13), Ablation: 24 h (n = 6), 48 h (n = 12), 72 h (n = 12). Two-tailed Student’s t tests were used. ns, p > 0.05; ****p < 0.0001.

Next, we closely monitored the spatial relationship between the calyx and the axon projections of the s-LNvs throughout different larval developmental stages (Figure 2A). We used OK107-GAL4 and Tab2-201Y-GAL4 to drive the expression of mCD8::GFP in all KC subtypes or specifically in γ KCs (Pauls et al., 2010), respectively, to visualize MB KCs. As development progressed, the calyx area defined by both GAL4 drivers exhibited a consistent expansion. Statistical analyses further confirmed that the calyx area increased proportionally with the advancement of developmental time, demonstrating a linear relationship (Figure 2B-C). Additionally, Pearson correlation analysis unveiled a robust positive correlation between the calyx area and the vertical projection distance of s-LNvs axon (r = 0.9987, P < 0.0001 for OK107-GAL4; r = 0.9963, P = 0.0003 for Tab2-201Y-GAL4) (Figure 2B-C). These compelling observations suggest that the MB calyx, and potentially specifically the calyx of γ KCs, exert significant influence on the vertical projection of the s-LNvs axon.

To validate our idea, we employed a straightforward approach to evaluate any changes in s-LNvs axon projection after ablating γ KCs using the essential Drosophila cell death activators Reaper (Rpr) and Head involution defective (Hid) driven by the Tab2-201Y GAL4. Initially, we confirmed the effectiveness of our approach. In the control group, the GFP signal in γ KCs remained unchanged, while the ablation group showed a significant reduction or complete absence of the signal (Figure 2-figure supplement 1). Subsequently, we examined the axon projection of s-LNvs under these conditions. In the non-ablated group, the vertical distance increased linearly during development, following the expected pattern (Figure 2D-E). However, in the ablation group, although the initial vertical projection of s-LNvs appeared normal, there was a noticeable decrease in vertical projection distance at 48 hours ALH. This vertical development trend ceased thereafter, with the vertical projection distance measuring only 20 to 30 µm at 72 hours ALH (Figure 2D-E). Importantly, in later larval stages, we observed that s-LNvs in certain larvae became undetectable (data not shown), suggesting the potential occurrence of apoptosis in these neurons following MB ablation. These results underscore a significant association between MB development, specifically the γ KCs, and the vertical projection dynamics of s-LNv axons.

s-LNv-expressed Dscam1 mediate s-LNv axons horizontal projection

To decipher the signaling pathways governing s-LNvs axon horizontal projection, we performed a screening with a total of 285 targets by using the RNA-interfering (RNAi) approach (Supplementary File 3). To optimize efficiency, we chose Clk856-GAL4, which is expressed in s-LNv from the embryonic stage (Figure 3-figure supplement 1A-B), although it is also expressed in other clock neuron subsets (Gummadova et al., 2009). Significantly, the knockdown of Down syndrome cell adhesion molecule (Dscam1) resulted in a notable decrease in the horizontal projection distance of s-LNv axons in late third instar larvae and adults (Figure 3-figure supplement 1C-F). In adults, knockdown of Dscam1-L (including the exon 19 isoform) in s-LNvs results in the failure of axon terminals to form normally (Zhang et al., 2019). To validate the role of s-LNv-expressed Dscam1 in governing axon horizontal projection, we employed the Pdf-GAL4, which specifically targeted s-LNvs during the larval stage. At 96 hours ALH, vertical projection showed no significant changes in Pdf-GAL4;UAS-Dscam1-RNAi flies, but a noticeable deficit was observed in the horizontal projection of the s-LNvs axon (Figure 3A-C). These observations strongly indicate that Dscam1 specifically regulates the horizontal projection of s-LNvs axon.

Critical role of Dscam1 in s-LNv axonal horizontal projection.

(A) Images of s-LNvs in Pdf >GFP and Pdf >GFP, Dscam1RNAi fly. White line segment represents the horizontal projection distance of s-LNvs. Larvae brains were stained with anti-PDF (white) and HRP (blue) antibodies at 96 hours ALH. Pdf >GFP: (n = 13), Pdf >GFP, Dscam1RNAi: (n = 15).

(B) Quantification of vertical projection length in Pdf >GFP and Pdf >GFP, Dscam1RNAi flies. Data are presented as mean ± SD. Two-tailed Student’s t tests, ns, p > 0.05.

(C) Quantification of horizontal A.U. in Pdf >GFP and Pdf >GFP, Dscam1RNAi flies. Data are presented as mean ± SD. Two-tailed Student’s t tests, ****p < 0.0001.

(D) Images of Dscam1 mutant s-LNvs projection phenotype. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (blue) antibodies.

(E) Quantification of horizontal A.U. in Dscam1 mutant flies. Data are presented as mean ± SD. w1118 (n = 22) Dscam121/+ (n = 5), Dscam11/+ (n = 8), Dscam105518/+ (n = 10), Dscam121/ Dscam105518(n = 6), Dscam121/Dscam11 (n = 10), Dscam105518(n = 14). One-way ANOVA with Tukey’s post hoc, ns, p > 0.05, ****p < 0.0001.

(F) Endogenous Dscam1 co-localizes with the s-LNvs axon terminal. Pdf >GFP fly heads were collected at 120 h ALH and stained with anti-Dscam1 (red). The white dotted line indicates the s-LNvs axon.

We further checked s-LNvs axon projection in several well-defined Dscam1 mutants, including Dscam121, Dscam11, and Dscam105518 (Hummel et al., 2003; Schmucker et al., 2000). While all heterozygous mutants displayed normal horizontal axon projection in s-LNvs, trans-heterozygous mutants and homozygotes exhibited shortened horizontal axon projection (Figure 3D-E). Additionally, our immunostaining results indicated concentrated Dscam1 expression at the growth cone of s-LNv axons in 3rd-instar larvae (Figure 3F). Taken together, these data demonstrate that Dscam1 controls the horizontal axon projection of s-LNvs in a cell-autonomous manner.

Dscam1 is a well-recognized cell adhesion molecule that plays a crucial role in axon guidance (Chen et al., 2006; Hummel et al., 2003; Zhan et al., 2004; Zhang et al., 2019). The receptors present on growth cones senses the extracellular axon guidance molecules to initiate the reorganization of the cellular cytoskeleton, leading to the facilitation of axonal projection. Convincingly, knockdown of the cytoskeletal molecules tsr (the Drosophila homolog of cofilin) (Sudarsanam et al., 2020) and chic (the Drosophila homolog of profilin) (Shields et al., 2014) in s-LNvs recaptured the horizontal axon projection deficits in Dscam1 knockdown flies or Dscam1 mutants (Figure 3-figure supplement 1C-D). Consistently, knockdown of Dock and Pak, the guidance receptor partners of Dscam1 (Schmucker et al., 2000), or SH3PX1, the critical linker between Dscam1 and the cytoskeleton (Worby et al., 2001), phenocopied Dscam1 knockdown flies or Dscam1 mutants (Figure 3-figure supplement 2A-B). Taken together, these findings provide solid evidence to support that Dscam1 signaling as a crucial regulator of s-LNvs axon horizontal projection (Figure 3-figure supplement 2C).

Neuronal-derived Netrins act upstream of Dscam1 to govern the horizontal projection of s-LNvs axon

Two classical guidance molecules, Slit and Netrin, have been shown to specifically bind to the extracellular domain of Dscam1 (Alavi et al., 2016; Andrews et al., 2008). The Drosophila genome contains the sole slit gene and two Netrin genes: Netrin-A (NetA) and Netrin-B (NetB) (Harris et al., 1996; Mitchell et al., 1996). Notably, ubiquitous knockdown of both NetA and NetB (hereafter referred to as Netrins), rather than slit, resulted in the defective horizontal axon projection of s-LNvs. Convincingly, NetA,NetB double mutant (Brankatschk & Dickson, 2006), but not NetA or NetB single mutant, exhibited the defective horizontal axon projection of s-LNvs (Figure 4A-B and Figure 4-figure supplement 1). These results imply that two Netrin molecules function redundantly in regulating horizontal axon projection of s-LNvs.

Neuron-derived Netrin guides s-LNvs horizontal projection.

(A) Images of immunostained in w1118, NetAΔ, NetBΔ and NetABΔ fly. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (blue) antibodies.

(B) Quantification of horizontal A.U. w1118, NetAΔ, NetBΔ and NetABΔ fly. Data are presented as mean ± SD, w1118 (n = 17), NetAΔ (n = 15), NetBΔ (n = 24), NetABΔ (n = 23). One-way ANOVA with Dunnett’s post hoc, ns, p > 0.05, ***p < 0.001.

(C) Images of immunostained in nSyb-GAL4 and repo-GAL4 knockdown Netrins. Larvae brains were collected at late 3rd larvae heads were stained with anti-PDF (white) and HRP (blue) antibodies.

(D) Quantification of horizontal A.U. in nSyb-GAL4 and repo-GAL4 knockdown Netrins fly. Data are presented as mean ± SD. nSyb-GAL4/+ (n = 13), nSyb-GAL4 >NetrinsRNAi (n = 16), repo-GAL4/+ (n = 34), repo-GAL4 >NetrinsRNAi (n = 20). Two-tailed Student’s t tests, ns, p > 0.05, ***p < 0.001.

To determine the source of Netrins, which are known to be secreted axon guidance molecules, we conducted experiments using pan-neuronal (nsyb-GAL4) and pan-glial (repo-GAL4) drivers to selectively knock down Netrins (Figure 4C-D). Strikingly, when Netrins were knocked down in neurons but not in glia, we observed severe defects in the horizontal axon projection of s-LNvs. These findings reveal that neuron-secreted Netrins serve as ligands for Dscam1, controlling the horizontal axon projection of s-LNvs.

Dorsal neuron-secreted Netrins mediate the horizontal axon projection of s-LNvs

To further identify the source of Netrins, we focused on the neurons located in the dorsolateral protocerebrum, where the s-LNvs axons change their projection direction. We first excluded the possibility that MB-secreted Netrins mediate the horizontal axon projection of s-LNvs, as knockdown of Netrins with OK107-GAL4 showed normal horizontal axon projection of s-LNvs (Figure 5A-B).

Netrin secreted by DN neurons guides time-specific horizontal projection of s-LNvs.

(A) Images of immunostained in OK107-GAL4 and per-GAL4, Pdf-GAL80 knockdown Netrins. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (blue) antibodies.

(B) Quantification of horizontal A.U. in OK107-GAL4 and per-GAL4, Pdf-GAL80 knockdown Netrins fly. Data are presented as mean ± SD. OK107-GAL4/+ (n = 20), OK107-GAL4 >NetrinsRNAi (n = 13). per-GAL4, Pdf-GAL80/+ (n = 18), per-GAL4, Pdf-GAL80 >NetrinsRNAi (n = 16), Two-tailed Student’s t tests for OK107-GAL4 knockdown Netrins, ns, p > 0.05. Mann-Whitney test for per-GAL4, Pdf-GAL80 knockdown Netrins, ****p < 0.0001.

(C) Images of DN ablation in the developing larval stages. Larvae brains were stained with anti-PDF (white) and HRP (blue) antibodies. Different time indicated hours ALH.

(D) Quantification of horizontal A.U. in Control (per-GAL4, Pdf-GAL80 >GFP) and Ablation (per-GAL4, Pdf-GAL80 >GFP,rpr,hid) flies. Data are presented as mean ± SD. Control: 72 h (n = 5), 96 h (n = 12), Ablation: 72 h (n = 5), 96 h (n = 10). Two-tailed Student’s t tests were used to compare conditions. ns, p > 0.05, ****p < 0.0001.

(E) Images of immunostained in per-GAL4, Pdf-GAL80 overexpress NetB in NetABΔ. Larvae brains were collected at late 3rd larvae. Heads were stained with anti-PDF (white) and HRP (blue) antibodies.

(F) Quantification of horizontal A.U. in per-GAL4, Pdf-GAL80 overexpress NetB in NetABΔ. Data are presented as mean ± SD. NetABΔ, UAS-NetB (n = 12), NetABΔ, per-GAL4,Pdf-GAL80 /+ (n = 5), NetABΔ,per-GAL4,Pdf-GAL80 >UAS-NetB (n = 9). One-way ANOVA with Bonferroni post hoc, ns, p > 0.05, ***p < 0.001.

The Dorsal neurons (DNs), a subset of clock neurons, also situate in the dorsolateral protocerebrum. DNs are categorized into three types, DN1, DN2, and DN3 (Reinhard et al., 2022), and previous studies have shown that both DN2 and DN1 form synaptic connections with s-LNvs axon terminals at the adult stage (Schlichting et al., 2022). Interestingly, knockdown of Netrins in a substantial portion of DN2, DN3, and DN1 (Kaneko et al., 1997) (Per-GAL4, Pdf-GAL80;UAS-Netrins-RNAi) resulted in the defective horizontal axon projection of s-LNvs (Figure 5A-B). To further validate the involvement of DNs in mediating the horizontal axon projection of s-LNvs, we conducted cell ablation experiments. Notably, the impairments in the horizontal axonal projection of s-LNvs due to DNs ablation were exclusively observed at 96 hours ALH, which is after the occurrence of horizontal projection (Figure 5C-D). In contrast, ablating crz+ neurons, which occupy a similar location to DNs, had no significant effects on the horizontal axon projection of the s-LNvs (Figure 5-figure supplement 1). Moreover, by expressing NetB in DNs in NetA,NetB double mutants, we successfully restored the defective horizontal axon projection of s-LNvs (NetABΔ;Per-GAL4, Pdf-GAL80;UAS-NetB) (Figure 5E-F). These results demonstrate that DNs secrete Netrins to guide the horizontal axon projection of s-LNvs.

Newborn DNs secrete Netrins to regulate the horizontal axon projection of s-LNvs

Finally, we wonder which population of DNs secrete Netrins to regulate the horizontal axon projection of s-LNvs. Therefore, we monitored the number and location of DNs during the s-LNvs axon projection and found that the number of DNs significantly increased during this process. Per-GAL4, Pdf-GAL80; UAS-mCD8::GFP only labeled 4-5 cells at 48 hours ALH, and raise to 10-15 by 72 hours, and subsequently increased to approximately 25 by 96 hours ALH (Figure 6A-B). It is worth noting that the location of these newly formed DNs resides lateral to the transition point of axon projection direction of the s-LNvs, while maintaining a basic parallelism with the horizontal axon projection of the s-LNvs (Figure 6A). The sharp increase in the number of DNs coincides remarkably with the timing of the switch in axon projection direction of the s-LNvs.

Dynamic changes in DN neurons coordinated with s-LNvs axonal targeting.

(A) Top: Images of s-LNvs axon and DNs growth process in the developing larval stages. Larvae brains were stained with anti-PDF (white) and GFP (green) antibodies. Different time indicated hours ALH are shown. Bottom: schematic diagram of s-LNvs horizontal projection and the corresponding increase in the number of DN neurons. s-LNvs axon (gray), DN neurons (green).

(B) Quantification of the number of DN neurons labeled by per-GAL4, Pdf-GAL80 at different developmental stages. Data are presented as mean ± SD.

(C) Images of newborn DN neurons were colocalized with Netrin-B at different developmental times. Larvae brains were stained with anti-GFP (green), mcherry (red) and PDF (white) antibodies. Different time indicated hours ALH. The white arrow demarcates the co-localization of red and green signals.

(D) Schematic representation s-LNvs projection directional transition and the corresponding increase in the number of DN neurons. s-LNvs axon (gray), DN neurons (green), newborn DN neurons (light green at 72 h, orange at 96 h).

Next, we asked whether these newly generated DNs expressed Netrins. To visualize the expression of endogenous NetB, we uesd the fly strains that insert either GFP or myc tag in NetB gene. At 72 hours ALH, we were unable to detect any NetB signals within the DNs. In contrast, at 96 hours ALH, we easily detected prominent NetB signals in approximately 6-8 newborn DNs (Figure 6C-D, Figure 6-figure supplement 1A).

Previous studies have shown that DSCAM is involved in Netrin-1-mediated axonal attraction (Andrews et al., 2008; G. Liu et al., 2009; Ly et al., 2008). In addition, DSCAM also functions as a repulsive receptor, associating with Uncoordinated-5C (UNC5C) to mediate Netrin-1-induced axon growth cone collapse (Fernandez et al., 2020; Purohit et al., 2012). To dissect the role of Netrin signaling, we ectopically expressed Netrins in neurons marked by R78G02-GAL4 (Jenett et al., 2012; Suzuki et al., 2022), located in front of the horizontal projection of s-LNv axons, closer to the midline (Figure 6-figure supplement 1B). Expression of either NetA or NetB in R78G02-GAL4-labeled neurons significantly suppressed the horizontal axon projection of s-LNvs (Figure 6-figure supplement 1B-C). Taken together, these findings reveal that newborn DNs secrete Netrins to orchestrate the transition of axon projection in s-LNvs from a vertical to a horizontal direction (Figure 7).

Dynamic cellular molecular environment during the s-LNv projection directional shift.

Cartoon depicting the dynamic cellular molecular microenvironment during the axonal projection directional shift from vertical to horizontal projection in s-LNvs. 72 h and 96 h represent 72 h and 96 h after larval hatching, respectively. s-LNvs (blue), neuropile (pink), brain lobe (gray circule), DNs (orange and light orange circles), optic lobe (gray dotted line), Netrin (Orange combination molding), Dscam1 (pruple).

Discussion

The mechanisms underlying axonal responses to intricate microenvironments and the precise development of axons within the brain have long remained enigmatic. In the past, researchers have identified a large number of axon guidance cues and receptors using different models in vivo and in vitro (Esther T Stoeckli, 2018). In addition, transient cell-cell interactions through intermediate targets play a crucial role in guiding axonal projection step by step towards its final destination (Chao et al., 2009; Garel & Rubenstein, 2004). The landmark midline, a crucial intermediate target, serves as a model in the majority of early studies exploring the directional transitions of axonal projections (Evans & Bashaw, 2010). However, so many neural projections do not cross the midline that it is challenging to understand how these axon projections are guided (Goodhill, 2016). In our study, we established an excellent model to investigate the mechanism of axonal projection direction switch by the Drosophila s-LNvs. We discovered a coordinated growth pattern between the vertical projection of s-LNvs and the MB calyx. Furthermore, the complementary interaction between Dscam1 expressed in s-LNvs and the emerging DN-secreted Netrins precisely modulates the transition of s-LNv axons from a vertical to a horizontal projection within a specific time window (Figure 7). These findings reveal the mechanism behind the transition of axonal projection direction, emphasizing the significance of developmental microenvironments in ensuring precise axon projection.

The dependence of s-LNv vertical projection on mushroom body calyx expansion

During our monitoring of the axonal projection of s-LNvs, we observed that the axon terminals reached the dorsolateral brain area at an early stage but continued their vertical growth. This phenomenon raises the question of what drives this growth. The mushroom body, a sophisticated central hub within the fruit fly’s brain, exhibits proximity to the projection of s-LNv axons in spatial arrangement. Upon careful examination, we found a positive correlation between the vertical length increase of s-LNv axons and the growth of the mushroom body calyx (Figure 2). Remarkably, when we selectively removed KC cells, we observed a striking effect on the axonal projection of s-LNvs. The s-LNv axonal projections either stalled at the initial stage or even completely disappeared (Figure 2 and data not shown). Hence, the vertical projection of s-LNv axons is dependent on mushroom body calyx expansion. Our hypothesis is supported by the findings of Helfrich-Förster et al., who reported locomotor activity and circadian rhythm defects in some MB mutants (HELFRICH-FÖRSTER et al., 2002), potentially attributed to the abnormality of s-LNvs axon. Moreover, it was observed that the MB mutants had minimal impact on the l-LNvs. This can be attributed to the fact that l-LNvs fibers emerge at a later stage during pupal development and are situated at a spatial distance from the mushroom body.

Unfortunately, we did not identify any molecules that have a discernible impact on the vertical projection of s-LNv axons through screening. This indicates that the successful completion of the vertical projection process may play a pivotal role in determining the overall existence of s-LNv axonal projections. Furthermore, the dorsal projection of s-LNv axons intersects with the ventral projection of neurons such as DN1 in the dorsolateral region adjacent to the calyx (Keene et al., 2011), resembling the corpus callosum in mammals (Fothergill et al., 2014; Hutchins et al., 2011; Keeble et al., 2006; Piper et al., 2009; Unni et al., 2012). Thus, during early developmental stages, multiple guiding cues may redundantly function between the mushroom body calyx and neighboring neural processes to ensure the smooth progression of neural development and establish a stable brain structure. However, to gain a more comprehensive understanding, it is worthwhile to explore whether the development of the mushroom body calyx influences the projection of other neurons traversing the same territory and the guidance cues involved. Insights from existing single-cell sequencing data obtained from multiple stages of larval life (Avalos et al., 2019; Corrales et al., 2022), might offer some clues and indications.

Netrin/Dscam signaling specifically controls the horizontal projection of s-LNv axons

In this paper, when we specifically knockdown Dscam1 in s-LNvs, we observed a significant defect in the axonal horizontal projection. This suggests that Dscam1 autonomously regulates the horizontal axonal projection of s-LNvs (Figure 3). Our findings align with a recent study that reported abnormal horizontal axonal projection of s-LNv neurons in adult flies when Dscam1 mRNAs lack the long 3’ UTR (Zhang et al., 2019).

The extracellular domain of Dscam1 has been verified to be capable of binding to Netrin or Slit (Alavi et al., 2016). In Drosophila, at the midline of the embryonic central nervous system, Dscam1 forms a complex with Robo1 to receive the Slit signal and promote the growth of longitudinal axons (Alavi et al., 2016). In adult flies, Slit/Robo signaling restricts the medial growth of s-LNv axons (Oliva et al., 2016). However, we found that reducing Netrins levels, rather than slit, led to a decrease in the horizontal projection distance of s-LNv axons. Moreover, in the NetABΔ double mutant, the extent of defects in the horizontal axonal projection of s-LNv neurons is similar to that observed when knocking down Dscam1 in s-LNvs. Indeed, previous studies have provided evidence showcasing the widespread involvement of Dscam1 as a receptor for Netrin in mediating axon growth and pathfinding (G. Liu et al., 2009; Ly et al., 2008; Matthews & Grueber, 2011). When we disturbed the expression of Netrin signals in the axon targeting microenvironment, it was also sufficient to cause abnormal s-LNv projection (Figure 6-figure supplement 1B-C). Therefore, our findings suggest that Netrin, as a ligand for Dscam1, regulates the process of switching axon projection direction (Figure 4).

The newborn DNs-secreted Netrin coordinates with s-LNv-expressed Dscam1 to switch the projection direction of s-LNv axons

While the midline serves as a prominent landmark, it presents a daunting challenge to comprehend the guidance and directional transitions of axons in many neural projections that do not actually cross this central axis (Evans & Bashaw, 2010). Segmented patterns such as the limb segment traversed by Ti1 neuron projections or neural circuits formed in a layer- or column-specific manner also serve as intrinsic “guideposts” (Bentley & Caudy, 1983; Isbister et al., 1999; A L Kolodkin et al., 1992), offering valuable insights into axonal pathfinding processes. How axonal projections without clear landmarks are guided? Following vertical projection, the s-LNvs growth cones remain within the dorsolateral area for at least 48 hours before initiating horizontal projection (Figure 1). This phenomenon aligns harmoniously with the outcomes observed in earlier studies that the axonal growth of other types of neurons slows down at the specific choice point, the midline (Bak & Fraser, 2003; Godement et al., 1994; Li et al., 2021), indicating that the dorsolateral area serves as an intermediate targets to facilitate the subsequent phase of s-LNvs projection journey.

Extracellular cues released by the final or intermediate targets play a crucial role in guiding axonal projection. The axon terminals of s-LNvs showed close spatial associations with the somas and processes of DNs. In this study, the knockdown of Netrins specifically in per+,Pdf- neurons, but not in the mushroom body, resulted in axon projection defects in s-LNvs (Figure 5). Furthermore, the ablation of DNs resulted in the inhibition of horizontal growth of s-LNvs axons (Figure 5). These findings suggest that the Netrin signaling microenvironment secreted by DNs is involved in regulating the horizontal projection of s-LNv axons.

Ideas and Speculation

The emergence of NetB-positive DNs and the aberrant axonal projection of s-LNv neurons caused by DN ablation occur concurrently during development (Figure 6). In Drosophila, three types of circadian oscillator neurons, including two DN1s, two DN2s, and 4-5 s-LNvs, can be identified as early as embryonic stage 16 (Houl et al., 2008). The DN3 group, which represents the largest contingent within the central circadian neuron network with over 35 neurons, emerges during the larval stage (T. Liu et al., 2015). Little is known about the function of DN3 neurons, particularly during the larval stage. Due to the lack of cell-specific labeling tools, we can only speculate about the identity of these newly generated Netrin-secreting neurons as DN3s based on their spatial distribution. Taken together, these results unveil a novel regulatory mechanism where axons, during the process of pathfinding, await guidance cues from newly born guidepost cells. This enables a switch in the direction of axon projection, facilitating the axon’s subsequent journey.

Supporting information